Text embedding models enable high-resolution insights into conceptual knowledge and learning in classroom-like settings

### 2023-01-31 paper/main.tex
finished fig 3 section

\begin{figure}[tp]
    \centering
    \includegraphics[width=0.7\textwidth]{figs/active-topics}

\caption{\textbf{Lecture and question topic overlap.} The bar plots display the
variability in topic weights across lecture timepoints (top panels) and
questions (bottom panels); colors denote topics. The top-weighted words from
the most ``expressive'' (i.e., variable across observations) topic from each
lecture are displayed in the upper right (orange: topic 2; yellow-green: topic
4). The top-weighted words from the full set of topics may be found in
Table~\topics.}

@@

We hypothesized that our text embedding model, which we had trained on
transcripts of the two lectures to characterize their conceptual contents,
should also allow us to capture the conceptual knowledge probed by each quiz
question. If our model successfully represented information about the deeper
conceptual content of the lectures (i.e., beyond surface-level details such
as particular word choices) then we should be able to recover a
correspondence between their embeddings and embeddings of other text that
reflect related concepts (e.g., quiz questions that were ostensibly, by
design, ``about'' one of the two lecture) despite differences in exact
verbiage. Intuitively, while the content of each given lecture will naturally
vary from moment to moment, the particular combination of dimensions along
which its trajectory varies across moments correspond to topics that encode
meaningful ``information''~\citep{Fish22}  about that lecture's dynamic
conceptual content. We therefore expected that quiz questions pertaining to
those same concepts might express the same set of topics.

%Figure~\ref{fig:topics} displays the variability in topic weights across timepoints in each lecture, and across questions in each category (i.e., lecture 1-related, lecture 2-related, and general physics knowledge).

We used our model to transformed the text of each question in our pool into
the same embedding space as the two lectures' trajectories (Fig
.~\ref{fig:sliding-windows}C). This yielded a single 15-dimensional
coordinate (i.e., topic vector) for each question. When we compared the
variability in topic weights across each lecture's timepoints, and across
questions from each category (i.e., lecture 1-related, lecture 2-related, and
general physics knowledge), we found multiple interesting patterns. First,
there is a strong correspondence between the sets of topics expressed by each
lecture and its relevant questions. This indicates that our model is
sensitive to the underlying conceptual content tested by different question,
and should allow us to differentially assess participants' knowledge for the
two lectures based on the \textit{specific} questions they answered correctly
or incorrectly on each quiz. Second, the topics expressed by the general
physics knowledge questions comprise a nonspecific mix of topics from the two
lectures (4 from lecture 1; 5 from lecture 2), most of which are expressed to
a relatively low degree. This indicates that our model should enable us to
discriminate questions that are highly relevant to our assessment of
participants' knowledge about concepts from the lectures from those that are
not (or potentially less so). Third, the two lectures' trajectories vary
along largely non-overlapping topic dimensions (e.g., variations in weight
for topics 2, 3, 8, 12, 13, and 15 are specific to lecture 1; variations in
weight for topics 1, 4, 5, 6, 7, and 10 are specific to lecture 2), but also
share a small number of expressed topics in common (topics 11 and 14). This 
indicates that our lecture is able to simultaneously capture

\begin{figure}[tp]
    \centering
    \includegraphics[width=\textwidth]{figs/lecture-question-similarity}

    \caption{\textbf{Which parts of each lecture are captured by each
    question?} Each panel displays timeseries plots showing how each question's
    topic vector correlates with each video timepoint's topic vector (Panel
    \textbf{A.}: correlations for the \textit{Four Fundamental Forces} lecture
    and associated questions; Panel \textbf{B.}: correlations for the
    \textit{Birth of Stars} lecture and associated questions). The colors denote
    question identities. The diamonds in each panel denote the moment of peak
    correlation between the indicated questions, in the indicated lectures. The
    associated questions' text, and snippets of the lectures' transcripts in
    the best-matching sliding windows, are displayed at the bottom of the
    figure.}

@@

Although a single lecture may be organized around a single broad theme at a
coarse scale, at a finer scale each moment of a lecture typically covers a
narrower range of content. We wondered whether a text embedding model trained
on the lectures' transcripts might capture some of this finer scale content.
For example, if a particular question asks about the content from one small
part of a lecture, we wondered whether our text embedding model could be used
to automatically identify the ``matching'' moment(s) in the lecture. When we
correlated each question's topic vector with the topic vectors for each second
of the lectures, we found some evidence that each question is temporally
specific (Fig.~\ref{fig:question-correlations}). In particular, most questions'
topic vectors were maximally correlated with a well-defined (and relatively
narrow) range of timepoints from their corresponding lectures, and the
correlations fell off sharply outside of that range. We also examined the
best-matching intervals for each question qualitatively by comparing the text
of the question to the text of the most-correlated parts of the lectures.
Despite that the questions were excluded from the text embedding model's
training set, in general we found (through manual inspection) a close
correspondence between the conceptual content that each question covered and
the content covered by the best-matching moments of the lectures. Two
representative examples are shown at the bottom of
Figure~\ref{fig:question-correlations}.

>>>>

We hypothesized that our text embedding model, which we had trained on
transcripts of the two lectures to characterize their conceptual contents,
should also allow us to capture the conceptual knowledge probed by each quiz
question. If our model successfully represented information about the deeper
conceptual content of the lectures (i.e., beyond surface-level details such
as particular word choices) then we should be able to recover a
correspondence between their embeddings and embeddings of other text that
reflects related concepts (e.g., quiz questions that were ostensibly, by
design, ``about'' one of the two lecture) despite differences in exact
verbiage. Intuitively, while the content of each given lecture will naturally
vary from moment to moment, the particular combination of dimensions along
which its trajectory varies across moments correspond to topics that encode
meaningful ``information''~\citep{Fish22}  about that lecture's dynamic
conceptual content. We therefore expected that quiz questions pertaining to
those same concepts might express the same set of topics.

\begin{figure}[tp]
    \centering
    \includegraphics[width=\textwidth]{figs/active-topics}

\caption{\textbf{Lecture and question topic overlap.} \textbf{A. Topic 
weight variability}. The bar plots display the variance of each topic's weight 
across lecture timepoints (top row) and questions (bottom row); colors 
denote topics. The top-weighted words from the most ``expressive'' (i.e., 
variable across observations) topic from each lecture are displayed in the 
upper right (orange: topic 2; yellow-green: topic 4). The top-weighted words 
from the full set of topics may be found in Table~\topics. \textbf{B. Relationships between topic weight variability.} Pairwise correlations between the distributions of topic weight variance for each lecture and question category. Each row/column corresponds to a bar plot in panel A.}

@@

We used our model to transform the text of each question in our pool into
the same embedding space as the two lectures' trajectories 
(Fig.~\ref{fig:sliding-windows}C). This yielded a single 15-dimensional
coordinate (i.e., topic vector) for each question. We then computed the variance
of each topic's weight across timepoints of each lecture, and across 
questions from each category (i.e., lecture 1-related, lecture 2-related, and
general physics knowledge; Fig.~\ref{fig:topics}A). Visual inspection of 
Figure~\ref{fig:topics}A reveals a strong correspondence between the sets of 
topics expressed by each lecture and its related questions, as well as limited 
overlap in the topics expressed by non-matched lecture-question set pairs. 
To quantify these apparent similarities and differences, we computed the correlation 
between each pair of topic-weight variance distributions (Fig.~\ref{fig:topics}B). 
We found that our model captured the contents of lecture-related questions 
using a weighted blend of topic dimensions that was highly similar to that of 
their reference lecture (\textit{Four Fundamental Forces} (\textit{FFF}) 
questions vs.~lecture: Pearson's $r(13) = 
0.824,~p<0.001,~95\%~\mathrm{confidence~interval~(CI)} = [0.696, 0.973]$; 
\textit{Birth of Stars} (\textit{BoS}) questions vs.~lecture: $r(13) = 
0.801,~p<0.001,~95\%~\mathrm{CI} = [0.539, 0.958]$) but diverged significantly 
from that of the non-reference lecture (\textit{FFF} questions vs.~\textit{BoS} 
lecture: $r(13) = -0.620,~p = 0.014,~95\%~\mathrm{CI} = [-0.871, -0.326]$; 
\textit{BoS} questions vs.~\textit{FFF} lecture: $r(13) = -0.550,~p = 
0.034,~95\%~\mathrm{CI} = [-0.803, -0.246]$). This indicated that our model 
was sufficiently sensitive to the conceptual content tested by different quiz questions to differentiate between those relating to one lecture versus the other, and allow us to

sensitive to the underlying conceptual 
content tested by different question,and should allow us to differentially assess 
participants' knowledge for the two lectures based on the \textit{specific}
questions they answered correctly or incorrectly on each quiz.
%When we compared the
%variability in topic weights across each lecture's timepoints, and across
%questions from each category (i.e., lecture 1-related, lecture 2-related, and
%general physics knowledge), we found multiple interesting patterns (Fig.~\ref{fig:topics}). First,
%there is a strong correspondence between the sets of topics expressed by each
%lecture and its relevant questions. This indicates that our model is
%sensitive to the underlying conceptual content tested by different question,
%and should allow us to differentially assess participants' knowledge for the
%two lectures based on the \textit{specific} questions they answered correctly
%or incorrectly on each quiz. Second, the topics expressed by the general
%physics knowledge questions comprise a nonspecific mix of topics from the two
%lectures (4 from lecture 1; 5 from lecture 2), most of which are expressed to
%a relatively low degree. This indicates that our model should enable us to
%discriminate questions that are highly relevant to our assessment of
%participants' knowledge about concepts from the lectures from those that are
%not (or potentially less so). Third, the two lectures' trajectories vary
%along largely non-overlapping topic dimensions (e.g., variations in weight
%for topics 2, 3, 8, 12, 13, and 15 are specific to lecture 1; variations in
%weight for topics 1, 4, 5, 6, 7, and 10 are specific to lecture 2), but also
%share a small number of expressed topics in common (topics 11 and 14). This
%indicates that our lecture is able to simultaneously capture

\begin{figure}[tp]
    \centering
    \includegraphics[width=\textwidth]{figs/lecture-question-similarity}
    \caption{\textbf{Which parts of each lecture are captured by each
    question?} Each panel displays timeseries plots showing how each question's
    topic vector correlates with each video timepoint's topic vector (Panel
    \textbf{A.}: correlations for the \textit{Four Fundamental Forces} lecture
    and associated questions; Panel \textbf{B.}: correlations for the
    \textit{Birth of Stars} lecture and associated questions). The colors denote
    question identities. The diamonds in each panel denote the moment of peak
    correlation between the indicated questions, in the indicated lectures. The
    associated questions' text, and snippets of the lectures' transcripts in
    the best-matching sliding windows, are displayed at the bottom of the
    figure.}

@@

Although an individual lecture may be organized around a single broad theme at a
coarse scale, at a finer scale each moment of a lecture typically covers a
narrower range of content. We wondered whether a text embedding model trained
on the lectures' transcripts might capture some of this finer scale content.
For example, if a particular question asks about the content from one small
part of a lecture, we wondered whether our text embedding model could be used
to automatically identify the ``matching'' moment(s) in the lecture. When we
correlated each question's topic vector with the topic vectors for each second
of the lectures, we found some evidence that each question is temporally
specific (Fig.~\ref{fig:question-correlations}). In particular, most questions'
topic vectors were maximally correlated with a well-defined (and relatively
narrow) range of timepoints from their corresponding lectures, and the
correlations fell off sharply outside of that range. We also examined the
best-matching intervals for each question qualitatively by comparing the text
of the question to the text of the most-correlated parts of the lectures.
Despite that the questions were excluded from the text embedding model's
training set, in general we found (through manual inspection) a close
correspondence between the conceptual content that each question covered and
the content covered by the best-matching moments of the lectures. Two
representative examples are shown at the bottom of
Figure~\ref{fig:question-correlations}.
